---
title: "Building Intuition for Interaction Terms (Solutions)"
output: pdf_document
params:
  p1q1_1: NA
  p1q1_2: NA
  p1q2: NA
  p1q3: NA
  p1q5: NA
  p1q7: NA
  p1q8: NA
  p1q9: NA
  p2q1: NA
  lungs: NA
  math: NA
---

```{r, include=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```

**Part 1**: The Lung Health Study (LHS) was a multicenter randomized clinical trial in the 1980s and 1990s, investigating whether a smoking intervention and the use of an inhaled bronchodilator (Atrovent) would decrease the rate of decline in lung function over the five-year follow-up period. A total of 5,887 participants (smokers aged 35-59 years old) were recruited from 10 clinical centers in the US and Canada and were randomized into three treatment groups (smoking intervention plus bronchodilator, smoking intervention plus placebo, or no intervention). The full, de-identified dataset is available for download within the app.

In this activity, we will be looking exclusively at baseline measurements and demographics of the LHS participants. In particular, we will examine the relationship between body mass index (BMI) and lung function, as measured by the ratio of FEV1 (forced expiratory volume in 1 second) to FVC (forced vital capacity). Higher values of FEV1/FVC % indicate better functioning lungs.

\vspace{.5in}

**Question 1**: To examine the distribution of the lung function measure (FEV1/FVC %), a `r params$p1q1_1` should be used. To visualize the relationship between BMI and lung function, a `r params$p1q1_2` should be used.

**Question 2**: What can we learn from these plots?

*Your answer*: `r params$p1q2`

So, now we've examined the overall relationship between BMI and lung function. But maybe there's more to the story. In particular, does biological sex affect the relationship between these variables? Here's the scatterplot again, but with separate lines of best fit for males and females, to help investigate this question:

```{r, echo=FALSE, message=FALSE, warning=FALSE}
ggplot(data = params$lungs) +
  geom_point(aes(x = bmi, y = FEVFVC02),
             alpha = .3) +
  geom_smooth(aes(x = bmi, y = FEVFVC02, color = sex),
              se = FALSE,
              method = "lm") +
  scale_color_manual(values = c("M" = "#D55E00",
                                "F" = "#0072B2")) +
  labs(x = "BMI",
       y = "FEV1/FVC % at baseline",
       color = "Sex",
       title = "Lines of best fit for males and females") +
  theme_minimal()
```

**Question 3**: Based on this scatterplot, describe in words how the relationship between BMI and lung function differs for males and females.

*Your answer*: `r params$p1q3`

**Question 4**: Now, let's fit a linear regression model with main effects for BMI and sex, and an interaction between the two. Which of the following equations represents this interaction model?

Here's the output after fitting this model in R:

```{r, echo=FALSE, message=FALSE, warning=FALSE}
summary(lm(FEVFVC02 ~ bmi + sex + bmi:sex,
           data = params$lungs))
```

**Question 5**: Interpret each estimated coefficient from the above output, including the model intercept.

*Your answer*: `r params$p1q5`

**Question 6**: Which of the model terms have interpretations that are *not* meaningful for this data?

\newpage

**Part 2**: In this part of the activity, we will look at academic data from schools in Portugal. We have a dataset of 395 students at two different secondary schools in Portugal from the 2005-2006 school year. The data were collected by reviewing student records and presenting students with questionnaires, with the intent of showing how data mining can predict academic performance. The dataset includes several potential predictors as well as our variable of interest: their final math score at the end of the academic year. The dataset is available for download within the app.

We want to evaluate the effect of years in high school on final math score varies between students living in rural vs. urban areas. (Here, urban/rural refers to where a student lives, not where their school is located.)

First, some EDA:

```{r, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(data = params$math) +
  geom_histogram(aes(x = hs_yrs), bins = 8,
                     fill = "grey", col = "black") +
  labs(x = "Years in high school",
       y = "Count",
       title = "Distribution of years in high school") +
  theme_minimal()

ggplot(data = params$math) +
  geom_boxplot(aes(y = address,
                   x = G3,
                   fill = address)) +
  scale_fill_manual(values = c("Rural" = "#D55E00",
                               "Urban" = "#0072B2")) +
  labs(y = "Urban/rural status",
       x = "Final math score",
       title = "Distribution of final math score by urban/rural status") +
  theme_minimal() +
  theme(legend.position = "none")

set.seed(20020521)
ggplot(data = params$math) +
  geom_jitter(aes(x = hs_yrs,
                  y = G3),
              width = .5,
              height = .5) +
  geom_smooth(aes(x = hs_yrs,
                  y = G3,
                  color = address),
              method = "lm",
              se = FALSE) +
  labs(x = "Years in high school",
       y = "Final math score",
       color = "Urban/rural",
       title = "Lines of best fit for rural and urban students") +
  scale_color_manual(values = c("Rural" = "#D55E00",
                                "Urban" = "#0072B2")) +
  scale_y_continuous(breaks = seq(0, 20, 1)) +
  theme_minimal()
```

\vspace{.5in}

**Question 1**: What can we learn from these plots?

*Your answer*: `r params$p2q1`

Here's the equation for the interaction model we want to fit:

$$Y_i = \beta_0 + \beta_1 \text{Urban}_i + \beta_2 \text{HS Years}_i + \beta_3 \text{Urban}_i \times \text{HS Years}_i + \varepsilon_i$$

**Question 2**: Using the model equation, find values for the regression coefficients to match the lines of best fit that were already plotted. To help you with this process, two scatterplots are presented below. The plot on the left shows the lines of best fit (as shown earlier), and the plot on the right shows the lines based on the values you choose. Your goal is to get the plot on the right to match the plot on the left. (You can use decimals!)

After finding the right values, here's the output after fitting this model in R:

```{r, echo=FALSE, message=FALSE, warning=FALSE}
summary(lm(G3 ~ address + hs_yrs + address:hs_yrs,
           data = params$math))
```

**Question 3**: Now consider the interpretations of our model coefficients. Select which term(s) have interpretation(s) subject to each limitation.
